Some examples in extended logic counting classes

### 2017-01-09 LICS/beyond.tex
sc

Given a relation signature $\R$, the set of transitive $\qfo$ formulas ($\tqfo$-formulas) is defined as an extension of $\qfo$ with the formula
%\begin{multline}
%	\label{eq-def-tqso}
%	\alpha := \varphi \, \mid \, s \, \mid \, (\alpha \add \alpha) \, \mid\, (\alpha \mult \alpha) \, \mid \, 
%	\sa{x} \alpha \, \mid\, \
%	\pa{x} \alpha \, \mid \\ 
%	\sa{X} \alpha \, \mid \, \pa{X} \alpha \, \mid \, [\pth \psi(\bar{x}, \bar{X},\bar{y}, \bar{Y})],
%\end{multline}
$
[\pth \psi(\bar{x},\bar{y})]
$, where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, and $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 
Given an $\R$-structure $\A$ with domain $A$, define a (directed) graph $\cG_{\psi}(\A) = (N,E)$ such that $N = A^k$ and for every pair $\bar b, \bar c \in N$, it holds that $(\bar b, \bar c) \in E$ if, and only if, $\A \models \psi(\bar b, \bar c)$.
Similar than for Example~\ref{ex:count-path}, we can count the paths of length at most $|A^k|$ in $ \cG_{\psi}(\A)$ with the formula $\beta_{\psi(\bar{x},\bar{y})}(\x, \y, \t, g)$:
\begin{multline*}
(\psi(\bar{x},\bar{y}) + \sa{\z} g(\x,\z,\t)\cdot \psi(\z,\y)) \cdot \varphi_{\text{\rm first-lex}}(\t) \ +\\
\sa{\t'} \varphi_{\text{succ-lex}}(\t',\t) \cdot \left(\sa{\x'} \sa{\y'} g(\x',\y',\t') \right),
\end{multline*}
where $\varphi_{\text{\rm first-lex}}$ and $\varphi_{\text{succ-lex}}$ are $\fo$-formulas defining the first and successor predicates over tuples in $A^k$, following the lexicographic order induced by~$<$.
Then the semantics of the path operator can be defined by using the following definition of $[\pth \psi(\bar{x},\bar{y})]$ in $\rqfo$:
\[
[\pth \psi(\bar{x}, \bar{y})] \ :=\ \sa{\t} (\varphi_{\text{\rm first}}(\t) \wedge \clfp{\beta_{\psi(\bar{x},\bar{y})}(\x,\y,\t,g)}).
\]
In other words, $\sem{[\pth \psi(\bar{x}, \bar{y})]}(\A,v)$ counts the number of paths from $v(\bar x)$ to $v(\bar y)$ in the graph $\cG_{\psi}(\A)$ whose length is at most~$|A^k|$.
%As for the case of $\qso$, the logic $\tqso(\LL)$ is obtained by restricting $\varphi$ in \eqref{eq-def-tqso} to be a formula in $\LL$. Moreover, the logic $\tqfo$ is obtained by disallowing in \eqref{eq-def-tqso} formulas $\sa{X} \alpha$ and $\pa{X} \alpha$, and by only allowing  first-order free-variables in the formula $\psi$ used in $[\pth \psi]$ in \eqref{eq-def-tqso}. 
As it was previously said, the operator for counting paths is exactly what we need to capture $\shl$.
\begin{theorem} \label{tqfo-shl}
	$\tqfo(\fo)$ captures $\shl$ over ordered structures.
\end{theorem}
This last result 
perfectly 
illustrates the benefits of our logical framework for the development of descriptive complexity for counting %functions.
complexity classes.  
The distinction in the language between the Boolean and the quantitative level allows us to define operators at the later level that cannot be defined at the former. 
As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extending with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L \in \qfo(\tc)$, where $\chi_L(w) = 1$ if $w \in L$, and $\chi_L(w) = 0$ otherwise. Thus, if we assume that

$\qfo(\tc)$ can def
In fact, it is not clear how $\shl$ could be capture by 
to capture $\shl$ with $\qfo(\tc)$\footnote{$\tc$ corresponds to $\fo$-logic extended with transitive closure.}, or any other extension of $\qfo(\fo)$ at the boolean level.
In contrast, with $\rqfo$ one can naturally define the path operator at the quantitative level and easily capture $\shl$.
Therefore, Theorem~\ref{tqfo-shl} is good example of how to go beyond classical logic to have a better understanding of the expressivity of counting.

>>>>

Given a relation signature $\R$, the set of transitive $\qfo$ formulas ($\tqfo$-formulas) is defined as an extension of $\qfo$ with the formula
%\begin{multline}
%	\label{eq-def-tqso}
%	\alpha := \varphi \, \mid \, s \, \mid \, (\alpha \add \alpha) \, \mid\, (\alpha \mult \alpha) \, \mid \, 
%	\sa{x} \alpha \, \mid\, \
%	\pa{x} \alpha \, \mid \\ 
%	\sa{X} \alpha \, \mid \, \pa{X} \alpha \, \mid \, [\pth \psi(\bar{x}, \bar{X},\bar{y}, \bar{Y})],
%\end{multline}
$
[\pth \psi(\bar{x},\bar{y})]
$, where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, and $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 
Given an $\R$-structure $\A$ with domain $A$, define a (directed) graph $\cG_{\psi}(\A) = (N,E)$ such that $N = A^k$ and for every pair $\bar b, \bar c \in N$, it holds that $(\bar b, \bar c) \in E$ if, and only if, $\A \models \psi(\bar b, \bar c)$.
Similar than for Example~\ref{ex:count-path}, we can count the paths of length at most $|A^k|$ in $ \cG_{\psi}(\A)$ with the formula $\beta_{\psi(\bar{x},\bar{y})}(\x, \y, \t, g)$:
\begin{multline*}
(\psi(\bar{x},\bar{y}) + \sa{\z} g(\x,\z,\t)\cdot \psi(\z,\y)) \cdot \varphi_{\text{\rm first-lex}}(\t) \ +\\
\sa{\t'} \varphi_{\text{succ-lex}}(\t',\t) \cdot \left(\sa{\x'} \sa{\y'} g(\x',\y',\t') \right),
\end{multline*}
where $\varphi_{\text{\rm first-lex}}$ and $\varphi_{\text{succ-lex}}$ are $\fo$-formulas defining the first and successor predicates over tuples in $A^k$, following the lexicographic order induced by~$<$.
Then the semantics of the path operator can be defined by using the following definition of $[\pth \psi(\bar{x},\bar{y})]$ in $\rqfo$:
\[
[\pth \psi(\bar{x}, \bar{y})] \ :=\ \sa{\t} (\varphi_{\text{\rm first}}(\t) \wedge \clfp{\beta_{\psi(\bar{x},\bar{y})}(\x,\y,\t,g)}).
\]
In other words, $\sem{[\pth \psi(\bar{x}, \bar{y})]}(\A,v)$ counts the number of paths from $v(\bar x)$ to $v(\bar y)$ in the graph $\cG_{\psi}(\A)$ whose length is at most~$|A^k|$.
%As for the case of $\qso$, the logic $\tqso(\LL)$ is obtained by restricting $\varphi$ in \eqref{eq-def-tqso} to be a formula in $\LL$. Moreover, the logic $\tqfo$ is obtained by disallowing in \eqref{eq-def-tqso} formulas $\sa{X} \alpha$ and $\pa{X} \alpha$, and by only allowing  first-order free-variables in the formula $\psi$ used in $[\pth \psi]$ in \eqref{eq-def-tqso}. 
As it was previously said, the operator for counting paths is exactly what we need to capture $\shl$.
\begin{theorem} \label{tqfo-shl}
	$\tqfo(\fo)$ captures $\shl$ over ordered structures.
\end{theorem}
This last result 
perfectly 
illustrates the benefits of our logical framework for the development of descriptive complexity for counting %functions.
complexity classes.  
The distinction in the language between the Boolean and the quantitative level allows us to define operators at the later level that cannot be defined at the former. 
As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extended with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L$ is in $\qfo(\tc)$, where $\chi_L(x) = 1$ if $x \in L$, and $\chi_L(x) = 0$ otherwise. Thus, if we assume that $\qfo(\tc)$ captures $\shl$ (over ordered structures), then we have that $\chi_L \in \nlog$ for every $L \in \nlog$. This implies that $\nlog = \ulog$,\footnote{A decision language $L$ is in $\ulog$ is there exists a logarithmic-space NTM $M$ accepting $L$ and satisfying that $\tma_M(x) = 1$ for every $x \in L$.} and thus contradicts the widely believed assumption that $\ulog \subsetneq \nlog$.
%$\qfo(\tc)$ can def
%In fact, it is not clear how $\shl$ could be capture by 
%to capture $\shl$ with $\qfo(\tc)$\footnote{$\tc$ corresponds to $\fo$-logic extended with transitive closure.}, or any other extension of $\qfo(\fo)$ at the boolean level.
%In contrast, with $\rqfo$ one can naturally define the path operator at the quantitative level and easily capture $\shl$.
%Hence, Theorem~\ref{tqfo-shl} is good example of how to go beyond classical logic to have a better understanding of the expressivity of counting.
